\bf No More, No Less Than Sum of its Parts: Monoids and The Algebra of Graphics, Statistics, and Interaction

### 2024-10-23 paper.Rmd
Wrote some more

```{r playfair}
#| echo: false
#| fig-cap: "\\label{fig:playfair}Examples of three methods for dealing with bars that have split into child components (segments). 
#| Top row: dodging (left) and stacking/highlighting (right), both by William Playfair [-@playfair1801; -@playfair1822]. 
#| Bottom row: layering, created with `ggplot2` [@wickham2010; the source of the underlying data is the U.S. fuel economy data set, @mpg2024]."
#| message: false
#| warning: false

@@

1. **Be selected**
2. **Represent selection**
3. (both in a visually and statistically sound way).

@@

```{r magritte}
#| echo: false
#| fig-height: 2.5
#| fig-show: "hold"
#| fig-align: "center"
#| fig-cap: "The representation of a thing is not the thing itself. Left: The Treachery of Images by René Magritte [-@magritte1929] makes the point that a painting of a pipe is not a pipe. Right: similarly, an image filled with coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data."

@@

```{r sumprodmax}
#| echo: false
#| fig-height: 2.5
#| fig-cap: "\\label{fig:sumprodmax} Different operators can be used for stacking. The values `1.2`, `1.8`, `NA`, and `1.5` were assigned to a 2-by-2 table and stacked using three different operators. Left: stacking using the sum operator. Middle: stacking using the product operator (note that the y-axis starts at 1). Right: stacking using the max operator." 
#| message: false
#| warning: false

@@

```{r}
#| echo: false
#| fig-height: 2.5
#| fig-cap: "\\label{fig:convexhull}Convex hull can be used as a valid representation of single-group selection but not multi-group selection. The light polygons shows the convex hull of the whole dataset, the dark polygon shows the hull of one selection group, the white dashed line shows the hull of a second selection group, and the medium-brightness polygon shows the combined hull of the two selection groups. Points show individual cases. Left: since convex hull is a monoid, we can use it to show a single highlighted group. Right: because convex hull lacks an inverse, after combining two hulls, we cannot guarantee we will be able to recover either of the original hulls - we do not know how many interior points were \"forgotten\". The source of the underlying data is the Palmer penguins data set [@horst2020]." 
#| message: false
#| warning: false

@@

p1 <- p0 + 
  geom_polygon(data = chull_all, fill = cols[1]) +
  geom_polygon(data = chull1, fill = cols[3]) +
  geom_point(col = "grey20", size = 0.25) +
  labs(title = "Single group: Valid")

p2 <- p0 + 
  geom_polygon(data = chull_all, fill = cols[1]) +
  geom_polygon(data = chull3, fill = cols[2]) +
  geom_polygon(data = chull1, fill = cols[3]) +
  geom_point(col = "grey20", size = 0.25) +
  geom_polygon(data = chull2, lty = "dashed", 
               col = "white", fill = NA) +
  labs(title = "Multiple groups: Invalid")

>>>>

```{r playfair}
#| echo: false
#| fig-cap: "\\label{fig:playfair}Examples of three methods for dealing with bars that have split into child components (segments). 
#| Top row: dodging (left) and stacking/highlighting (right), both by William Playfair [-@playfair1801; -@playfair1822]. 
#| Bottom row: layering, created with `ggplot2` 
#| [@wickham2010; the source of the underlying data is the U.S. fuel economy data set, @mpg2024]."
#| message: false
#| warning: false

@@

1. Be selected
2. Represent selection
3. (both in a visually and statistically sound way).

@@

```{r}
#| echo: false
#| fig-cap: "\\label{fig:magritte}The representation of a thing is not the thing itself. 
#| Left: The Treachery of Images by René Magritte [-@magritte1929] makes the point that a painting of a pipe is not a pipe. 
#| Right: similarly, an image filled with coloured rectangles is not a barplot, 
#| unless the rectangles accurately represent some underlying data."
#| fig-height: 2.5
#| fig-show: "hold"

@@

```{r sumprodmax}
#| echo: false
#| fig-height: 2.5
#| fig-cap: "\\label{fig:sumprodmax}Different operators can be used for stacking. The values `1.2`, `1.8`, `NA`, and `1.5` were assigned to a 2-by-2 table and stacked using three different operators. Left: stacking using the sum operator. Middle: stacking using the product operator (note that the y-axis starts at 1). Right: stacking using the max operator." 
#| message: false
#| warning: false

@@

```{r}
#| echo: false
#| fig-height: 2.5
#| fig-cap: "\\label{fig:convexhull}Convex hull can be used as a valid representation of single-group selection but not multi-group selection. The light polygons shows the convex hull of the whole dataset, the dark polygon shows the hull of one selection group, the white dashed line shows the hull of a second selection group, and the medium-brightness polygon shows the combined hull of the two selection groups. Points show individual cases. Left: since convex hull is a monoid, we can use it to show a single highlighted group. Right: because convex hull lacks an inverse, after combining two hulls, we cannot guarantee that we will be able to recover either of the original hulls - we do not know how many interior points were \"forgotten\". The source of the underlying data is the Palmer penguins data set [@horst2020]." 
#| message: false
#| warning: false

@@

p1 <- p0 + 
  geom_polygon(data = chull_all, fill = cols[1]) +
  geom_polygon(data = chull1, fill = cols[3]) +
  geom_point(col = "grey20", size = 0.25) +
  labs(title = "Valid representation")

p2 <- p0 + 
  geom_polygon(data = chull_all, fill = cols[1]) +
  geom_polygon(data = chull3, fill = cols[2]) +
  geom_polygon(data = chull1, fill = cols[3]) +
  geom_point(col = "grey20", size = 0.25) +
  geom_polygon(data = chull2, lty = "dashed", 
               col = "white", fill = NA) +
  labs(title = "Not a valid representation")

### 2024-10-23 paper.tex
Wrote some more

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Be selected}
\item
  \textbf{Represent selection}
\item
  (both in a visually and statistically sound way).
\end{enumerate}

@@

\begin{figure}

{\centering \includegraphics{paper_files/figure-latex/magritte-1}

}

\caption{The representation of a thing is not the thing itself. Left: The Treachery of Images by René Magritte [-@magritte1929] makes the point that a painting of a pipe is not a pipe. Right: similarly, an image filled with coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data.}\label{fig:magritte}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/sumprodmax-1.pdf}
\caption{\label{fig:sumprodmax} Different operators can be used for
stacking. The values \texttt{1.2}, \texttt{1.8}, \texttt{NA}, and
\texttt{1.5} were assigned to a 2-by-2 table and stacked using three
different operators. Left: stacking using the sum operator. Middle:
stacking using the product operator (note that the y-axis starts at 1).
Right: stacking using the max operator.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/unnamed-chunk-2-1.pdf}
\caption{\label{fig:convexhull}Convex hull can be used as a valid
representation of single-group selection but not multi-group selection.
The light polygons shows the convex hull of the whole dataset, the dark
polygon shows the hull of one selection group, the white dashed line
shows the hull of a second selection group, and the medium-brightness
polygon shows the combined hull of the two selection groups. Points show
individual cases. Left: since convex hull is a monoid, we can use it to
show a single highlighted group. Right: because convex hull lacks an
inverse, after combining two hulls, we cannot guarantee we will be able
to recover either of the original hulls - we do not know how many
interior points were ``forgotten''. The source of the underlying data is
the Palmer penguins data set \citep{horst2020}.}
\end{figure}

>>>>

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Be selected
\item
  Represent selection
\item
  (both in a visually and statistically sound way).
\end{enumerate}

@@

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/unnamed-chunk-2-1.pdf}
\caption{\label{fig:magritte}The representation of a thing is not the
thing itself. Left: The Treachery of Images by René Magritte
\citeyearpar{magritte1929} makes the point that a painting of a pipe is
not a pipe. Right: similarly, an image filled with coloured rectangles
is not a barplot, unless the rectangles accurately represent some
underlying data.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/sumprodmax-1.pdf}
\caption{\label{fig:sumprodmax}Different operators can be used for
stacking. The values \texttt{1.2}, \texttt{1.8}, \texttt{NA}, and
\texttt{1.5} were assigned to a 2-by-2 table and stacked using three
different operators. Left: stacking using the sum operator. Middle:
stacking using the product operator (note that the y-axis starts at 1).
Right: stacking using the max operator.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{\label{fig:convexhull}Convex hull can be used as a valid
representation of single-group selection but not multi-group selection.
The light polygons shows the convex hull of the whole dataset, the dark
polygon shows the hull of one selection group, the white dashed line
shows the hull of a second selection group, and the medium-brightness
polygon shows the combined hull of the two selection groups. Points show
individual cases. Left: since convex hull is a monoid, we can use it to
show a single highlighted group. Right: because convex hull lacks an
inverse, after combining two hulls, we cannot guarantee that we will be
able to recover either of the original hulls - we do not know how many
interior points were ``forgotten''. The source of the underlying data is
the Palmer penguins data set \citep{horst2020}.}
\end{figure}
